---
title: Working with Characters
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`


```{r echo=FALSE}
library(stringr)
```

Working with character strings is one of the most common tasks in R. In this section we discuss some of the routines we have for that.

Character strings can use single or double quotes:

```{r}
'this is a string'
"this is a string"
```

Say you want to type in a vector of names. Having to type all those quotes is hazzle, but there is a nice routine in the *Hmisc* package that helps:

```{r}
library(Hmisc)
Cs(Joe, Jack, Ann, Laura)
```

`r hl()$hr()`

*Resma3* has the data set *agesexUS*, which has the breakdown of the US population by gender and age according to the 2000 US Census. We are going to work with the names of the states (plus DC and PR) for a bit:

```{r}
states <- agesexUS$State
head(states)
```

to find out how long a string is use

```{r}
nchar(states)
```

What state has the longest name?

```{r}
states[which(nchar(states) == max(nchar(states)))]
```

Say we want to shorten the strings to just the first three letters:

```{r}
substring(states, first=1, last=3)
```

Now, though, several of the strings are the same ("Ala"). If that is a problem use

```{r}
abbreviate(states)[1:6]
```

This routine figures out what the length of the shortest string is that makes all of them unique (here 4). We can make it a little longer if we want:


```{r}
abbreviate(states, minlength = 6)[1:6]
```


Notice that this keeps the full strings as names.

`r hl()$hr()`

Say we want the last 3 letters of the states names:

```{r}
substring(states, 
          first = nchar(states)-2,
          last = nchar(states))[1:6]
```

Let's say we want all the states whose name starts with P:

```{r}
grep("P", states)
```

tells us those are the states at position 39 and 52, so now

```{r}
states[grep("P", states)]
```

or directly:

```{r}
grep("P", states, value = TRUE)
```

Here is another way to do this:

```{r}
states[startsWith(states, "P")]
```

Very useful is its partner:

```{r}
states[endsWith(states, "o")]
```

This can be used for example to find the files of a certain type in a folder:

```{r}
dir()[endsWith(dir(), ".Rmd")][1:5]
```

`r hl()$vspace()`

Notice that above we only got the states whose names have a capital P. What if we want all states with either p or P?

```{r}
grep(pattern = "[pP]", x = states, value = TRUE)
```

the syntax [pP] matches either p or P. This is an example of a *regular expression*, which we will discuss shortly.

We can also use the function *tolower*, which turns all the letters into lower case:

```{r}
tolower(states)[1:4]
grep("p", tolower(states), value = TRUE)
```

but now all the letters are lower case.

Of course there is also a *toupper* function.

`r hl()$hr()`

We have the *grepl* function, which returns TRUE if the string contains the pattern:

```{r}
states[1:6]
grepl(pattern = "s", states)[1:6]
```

Suppose we want to replace all the A's with *'s:

```{r}
gsub("A", "*", states)[1:6]
```

There is also the *sub* function, which does the same as *gsub* but only to the first occurance:

```{r}
sub("a", "A", c("abba"))
gsub("a", "A", c("abba"))
```

**Exercise**
```{r echo=FALSE, comment=""}
y <- sample(letters, size=150, replace=TRUE)
```

```{r comment=""}
cat(y, "\n")
```

How many a' are in this string? 

```{r, echo=FALSE, eval=FALSE}
length(y[grep("a", y)])
```

You can get the vector into R as follows: copy it and in R run


```{r eval=FALSE}
# Windows
x <- scan("clipboard", what="char")
# Mac
x <- scan(pipe("pbpaste"), what="char")
```


`r hl()$hr()`

Let’s ask the following question: what is the distribution of the vowels in the names of the states?  For instance, let’s start with the number of a’s in each name.  There’s a very useful function for this purpose: *gregexpr*.  

We can use it to get the number of times that a searched pattern is found in a character vector.  When there is no match, we get a value -1.

```{r}
positions_a <- gregexpr(pattern = "a", 
                        text = states, 
                        ignore.case = TRUE)
positions_a[[1]]
```

tells us that in "Alabama" there are a's in positions 1, 3, 5 and 7.

Now we need to go through all the states names and find out how many a's there are in each. Here is a fast way to do this, using one of the *apply* functions:


```{r}
f <- function(x) {
  ifelse(x[1] > 0, length(x), 0)
# if there is no a, x is -1, so we get 0  
}
num_a <- sapply(positions_a, f)
num_a
```

Now let's do this for all the vowels:

```{r}
vowels <- c("a", "e", "i", "o", "u")
num_vowels <- rep(0, 5)
names(num_vowels) <- vowels
for(i in seq_along(vowels)) {
  positions <- gregexpr(pattern = vowels[i], 
                        text = states, 
                        ignore.case = TRUE)
  num_vowels[i] <- sum(sapply(positions, f))
}
num_vowels
```

`r hl()$hr()`

# paste

One of the most useful commands in R is *paste*. It let's us put together various parts as a string:

```{r}
paste(1:3)
paste("a", 1:3)
paste("a", 1:3, sep=":")
paste("a", 1:3, sep="")
paste0("a", 1:3)
```

Notice that the last two are the same.

If we want to make a single string use

```{r}
paste0("a", 1:3, collapse="")
paste0("a", 1:3, collapse="-")
```

**Exercise**

write a routine that generates a licence plate in Puerto Rico at random. For example

```{r echo=FALSE}
license.plate <- function()
  paste0(
    paste(sample(LETTERS, size=3, replace=TRUE),
          collapse=""),
    paste(sample(0:9, size=3, replace=TRUE),
          collapse="")
    )
```

```{r}
license.plate()
```

`r hl()$hr()`

paste "combines" stuff into a string. Sometimes we want to do the opposite:

```{r}
txt <- "This is a short sentence"
strsplit(txt, " ")
```

notice that the result is a list, so often we use
 
```{r}
unlist(strsplit(txt, " "))
```
 
# Regular Expressions

A *regular expression* (a.k.a. regex) is a special text string for describing a certain amount of text.  This “certain amount of text” receives the formal name of pattern.  Hence we say that a regular expression is a pattern that describes a set of strings. 

Tools for working with regular expressions can be found in virtually all scripting languages (e.g.  Perl,  Python,  Java,  Ruby,  etc). R has  some  functions  for  working  with  regular  expressions  although  it  does  not  provide  the  wide  range  of  capabilities  that  other  scripting languages do.  Nevertheless, they can take us very far with some workarounds (and a bit of patience).

To know more about regular expressions in general, you can find some useful information in the following resources:

-  Regex wikipedia [http://en.wikipedia.org/wiki/Regular_expression](http://en.wikipedia.org/wiki/Regular_expression)

-  Regular-Expressions.info website (by Jan Goyvaerts) [http://www.regular-expressions.info](http://www.regular-expressions.info)

The main purpose of working with regular expressions is to describe patterns that are used to match against text strings.  Simply put, working with regular expressions is nothing more than pattern matching.

The result of a match is either successful or not. The simplest version of pattern matching is to search for one occurrence (or all occurrences)
of some specific characters in a string.  For example, we might want to search for the word "programming" in a large text document, or we might want to search for all occurrences of the string "apply" in a series of files containing R scripts.

The most important use of regular expressions is in the replacement of a pattern, say using gsub. Regular expressions allow us to not just use specific characters as patterns but much more general things:

Let's take the vector

```{r echo=2}
txt <- c("In", 2017, "there", "where", 17, "hurricanes")
cat(txt)
```

Let's say I want to pick out those elements of the vector that are (or at least could be) numeric. Here is one way to do it:

```{r}
as.numeric(txt)
txt[!is.na(as.numeric(txt))]
```

or we can use regexp:

```{r}
txt[grepl("\\d", txt)]
```


say we want to replace the spaces in a sentence with the underscore:

```{r}
gsub("\\s", "_", "Not a very interesting sentence")
```

We already used [pP] before to match both small and large cap p's. This is in fact a regular expression:

```{r}
gsub("[0-9]", "%", txt)
```

A ^ in front is NOT:

```{r}
gsub("[^0-9]", "%", txt)
```


## POSIX

Closely related to the regex character classes we have what is known as
POSIX  character classes. In R, POSIX character classes are represented with expressions inside double brackets [[ ]]. 

[[:lower:]] Lower-case letters  
[[:upper:]] Upper-case letters
[[:alpha:]]  Alphabetic characters ([[:lower:]] and [[:upper:]])  
[[:digit:]] Digits:  0, 1, 2, 3, 4, 5, 6, 7, 8, 9  
[[:alnum:]] Alphanumeric characters ([[:alpha:]] and [[:digit:]])  
[[:blank:]] Blank characters:  space and tab 
[[:cntrl:]] Control characters  
[[:punct:]] Punctuation characters:  !  ” #  % & ’ ( ) * + , - .  / :  ;  
[[:space:]] Space characters:  tab, newline, vertical tab, form feed,
carriage return, and space  
[[:digit:]] Hexadecimal digits:  0-9 A B C D E F a b c d e f  
[[:print:]] Printable characters ([[:alpha:]], [[:punct:]] and space)
[[:graph:]] Graphical characters ([[:alpha:]] and [[:punct:]])

so we could also do this

```{r}
as.numeric(txt[grepl("[[:digit:]]", txt)])
```


# Some Examples

## Palindrome

a *palindrome* is a word that is the same when read forwards or backwards. Some examples are noon, civic, radar, level, rotor, kayak, reviver, racecar, redder, madam, and refer. Let's write a sequence of commands that take a sentence and return any palindromes. As an example, consider

```{r}
txt <- "At Noon the Meteorologist is checking the Radar"
```

which should result in the vector ("noon", "radar").

First we need to split the sentence into words:

```{r}
wrds <- unlist(strsplit(txt, " "))
wrds
```

Next we need to turn each word around:

```{r}
n <- length(wrds)
rev.wrds <- rep("", n)
for(i in 1:n) 
  rev.wrds[i] <- paste(unlist(strsplit(wrds[i], ""))[nchar(wrds[i]):1], collapse = "")
rev.wrds
```

Finally, let's check whether the words are the same, but taking into account that Noon is still a palindrome!

```{r}
wrds[tolower(wrds) == tolower(rev.wrds)]
```


### Email Addresses

Consider the web site of the Math department at [http://math.uprm.edu/academic/people.php](http://math.uprm.edu/academic/people.php)

Let's say we want to write a routine that picks out all the email addresses.

First we need to download the web site. This can be done with the scan command because as is explained in the help file the argument can be a *connection*, which includes URL's

```{r eval=FALSE}
txt <- scan("http://math.uprm.edu/academic/people.php", 
      what="char", sep="\n")
```

```{r echo=FALSE}
txt <- scan("C:\\Users\\Wolfgang\\Dropbox\\teaching\\Computing-with-R\\emails.txt", what="char", sep="\n")
```

Next we need to figure out what defines an email address. Obviously it needs to have the @ symbol, so let's go through the text and pick out those lines that have the @ symbol:

```{r}
sum(unlist(gregexpr("@", txt))>0)
txt <- grep("@", txt, value = TRUE)
length(txt)
```

So the @ symbol appears 236 time. but strangly there are only 5 lines with @ symbols! That is because 

```{r}
substring(txt[1], 1, 500)
```

so on the website the addresses are in an html table, which was read in as a single string. We can see that immediately after each email address is the text \</a\> , which is the html tag to end a link. Let's split up the text according to the \</a\> tag:

```{r}
txt <- unlist(strsplit(paste(txt, collapse=""), "</a>"))
txt[1:2]
```

Ok, next we have to eliminate all the lines that don't have a @ in it:

```{r}
txt <- grep("@", txt, value = TRUE)
length(txt)
```

which is good because 2*118=236, and each address appeared twice in the table. So we got all of them.

Finally we need to extract the email adress from each line. Checking them we see that just before each address is an empty space, so maybe this will work:

```{r}
txt <- unlist(str_split(txt, " "))
txt <- grep("@", txt, value = TRUE)
txt[1:4]
```

almost, we just need to get rid of those lines starting with href:

```{r echo=-1}
txt <- txt[-107]
txt <- txt[!grepl("href", txt)]
txt
```

If we wanted to send an email to all the people in the department we could now use the write command to copy them to the clipboard, switch to a mail program and copy them into the address box.

Programs like these are routinely used to go through millions of websites and search for email addresses, which are then sent spam emails. This is why I only write mine like this:

wolfgang[dot]rolke[at]upr[dot]edu

### Example: Binary Arithmetic

We have previously discussed binary arithmetic. There we used a simple vector of 0's and 1's. The main problem with that is that it is hard to vectorize the routines. Instead we will now use character sequences like "1001". 

We have previously written several functions for this. We will want to reuse them but also adapt them to this new format. To do so we need to turn a character string into a vector of numbers and vice versa. Also we want our routines to be vectorized:

-  Decimal to Binary

```{r}
decimal.2.binary <- function(x) {
  n <- length(x)
  y <- rep("0", n)
  for(k in 1:n) {
    if(x[k]==0 | x[k]==1) {  # simple cases 
        y[k] <- x[k]
        next
    }    
    i <- floor(log(x[k], base=2)) # largest power of 2 less than x
    bin.x <- rep(1, i+1) # we will need i+1 0'1 and 1's, first is 1
    x[k] <- x[k]-2^i 
    for(j in (i-1):0) {
       if(2^j>x[k]) 
         bin.x[j+1] <- 0
       else {
         bin.x[j+1] <- 1
         x[k] <- x[k]-2^j
       }
    }
    y[k] <- paste(bin.x[length(bin.x):1], collapse="")
  }
  y
}
decimal.2.binary(c(7, 8, 26))
```

-  Binary to Decimal

```{r}
binary.2.decimal <- function(x){
  n <- length(x)
  y <- rep(0, n)
  for(i in 1:n) {
    tmp <- as.numeric(strsplit(x[i], "")[[1]])
    y[i] <- sum(tmp*2^(length(tmp):1-1))
  }
  y
}  
binary.2.decimal(c("111", "1000", "11010"))
binary.2.decimal(decimal.2.binary(126))
decimal.2.binary(binary.2.decimal(c("100101")))
```

-  *is. binary*:

```{r}
is.binary <- function(x) {
  n <- length(x)
  y <- rep(TRUE, n)
  for(i in 1:n) {
      x.vec <- as.numeric(strsplit(x[i], "")[[1]])
      if(all(x.vec==0)) {
          y[i] <- TRUE 
          next
      }
      x.vec <- x.vec[x.vec!=0]
      x.vec <- x.vec[x.vec!=1]
      if(length(x.vec)==0) y[i] <- TRUE 
      else y[i] <- FALSE 
  }
  y
}  
is.binary(c("1001", "0", "11a1"))
```

-  addition

Here I am going to reuse the routine we had already written:

```{r}
binary_addition <- function(x, y) {
# First make x and y of equal length and with one extra 
# slot in case it's needed for carry over
# Fill x and y with 0's as needed. 
  n <- length(x)
  m <- length(y)
  N <- max(n, m)+1
  x <- c(rep(0, N-n), x)
  y <- c(rep(0, N-m), y)  
  s <- rep(0, N) #  for result
  ca <- 0 # for carry over term
  for(i in N:1) {
      n <- x[i]+y[i]+ca
      if(n<=1) {# no carry over
        s[i] <- n
        ca <- 0
      }  
      else {# with carry over
        s[i] <- 0
        ca <- 1
      }
    }
  if(s[1]==0) s <- s[-1]# leading 0 removed if necessary
  s
}
binary_addition(c(1, 0), c(1, 1, 0))
```



```{r}
binary.addition<- function(x, y) {
  n <- length(x)
  m <- length(y)
  if(m!=n) cat("Vectors have to have the same length!\n")
  s <- rep("0", n)
  for(i in 1:n) {
    x.vec <- as.numeric(strsplit(x[i], "")[[1]])
    y.vec <- as.numeric(strsplit(y[i], "")[[1]])
    tmp <- binary_addition(x.vec, y.vec)
    s[i] <- paste(tmp, collapse="")
  }
  s
}  
binary.addition(c("0", "10", "1001"), c("0", "110", "1101"))
```

Let's turn this into an infix addition operator:

```{r}
'%+b%' <- function(x, y) binary.addition(x, y)
x <- c("10", "1001", "100101", "11101001") 
y <- c("101", "1001", "10101", "1001100") 
binary.2.decimal(x)
binary.2.decimal(y)
z <- x %+b% y 
x
binary.2.decimal(z)
```

Let's define a new class of objects "binary numbers":

```{r}
as.binary <- function(x) {
  class(x) <- "binary"
  return(x)
}
```

what methods might be useful here? Let's write two:

-  print

this is how our number will appear when we use print(x):

```{r}
print <- function(x) UseMethod("print")
print.binary <- function(x) {
  n <- length(x)
  for(i in 1:length(x)) {
    y <- as.numeric(strsplit(x[i], "")[[1]])
    y <- paste(y, collapse = ".")
    cat(y, "\n")
  }

}
x <- as.binary(c("10", "1001", "100101"))
print(x)
```

- summary

what should we calculate as summary statistics? Let's do three:

  - how many  
  - most frequent (mode, NA if all only once)  
  - percentage of 0's  
  

```{r}
summary <- function(x) UseMethod("summary")
summary.binary <- function(x) {
  n <- length(x)
  if(length(unique(x))==length(x)) mode <- NA
  else {
    z <- table(x)
    z <- z[z==max(z)]
    mode <- names(z)
  }  
  y <- paste(x, collapse = "") # one long string
  y <- as.numeric(strsplit(y, "")[[1]]) # vector of 0's and 1's
  y <- round(sum(y==0)/length(y)*100, 1)
  cat("N =", n, "\n")
  cat("Mode =", mode, "\n")
  cat("% 0's =", y, "\n")
}
x <- sample(1:100, size=1000, replace=TRUE)
x <- as.binary(decimal.2.binary(x))
print(x[1:5])
summary(x)
```
  
  
