---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Data Formats: Vectors, Matrices, Arrays, Dataframes and Lists

##Vectors

here are some useful commands for vectors:

```{r}
x <- c(1, 2, 3, 4, 5, 6)
x
length(x)
names(x) <- LETTERS[1:6]
x
```

If we have several vectors of the same type and length which belong together we can put them in a 

##Matrix

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c(4, 2, 3, 5, 3, 4)
z <- c(3, 4, 2, 3, 4, 2)
cbind(x, y, x)
rbind(x, y, z)
```

here are several other ways to make a matrix:

```{r}
matrix(x, 2, 3)
matrix(x, ncol=2)
matrix(0, 3, 3)
diag(3)
```

we can control how the matrix is filled in:

```{r}
matrix(1:8, nrow=2)
matrix(1:8, nrow=2, byrow = TRUE)
```

just like vectors a matrix is all of the same type, with R doing type conversion if necessary:

```{r}
matrix(c(1, 2, "A", 3), 2, 2)
```

useful commands for R matrices are:

```{r}
x <- matrix(c(1, 2, 3, 4, 5, 6), 2, 3)
x
dim(x)
nrow(x)
ncol(x)
dimnames(x) <- list(c("A", "B"), c("a", "b", "c"))
x
colnames(x)
rownames(x)
colnames(x) <- c("Height", "Age 1", "Age 2")
x
```

Here is a strange way to make a matrix:

```{r}
x <- 1:6
x
dim(x) <- c(2, 3)
x
```


What happens if we try to make a matrix out of a vector that doesn't have the right number of entries?

```{r warning=TRUE}
matrix(1:5, 2, 2)
```

not surprisingly it just uses the elements needed.

```{r warning=TRUE}
matrix(1:3, 2, 2)
```

In this case R just starts all over again. This is called *recycling*.

In either case R gives a warning.

##Arrays

an array is a k-dimensional matrix. For example

```{r}
array(1:8, dim=c(2, 2, 2))
```

Often these come about as a 3-way table: 

```{r}
x <- sample(1:3, size=20, replace=T)
y <- sample(c("F", "M"), size=20, replace = T)
z <- sample(c("a", "b", "c"), size=20, replace = T)
table(x,y,z)
```


##Dataframe

sometimes we have several vectors of the same length but of different types, then we can put them together as a data frame:

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c("a", "a", "b", "c", "a", "c")
z <- c(T, T, F, T, F, T)
xyz <- data.frame(x, y, z)
xyz
```

the same commands as for matrices work for data frames as well:

```{r}
dim(xyz)
nrow(xyz)
ncol(xyz)
dimnames(xyz) <- list(letters[1:6], c("a", "b", "c"))
xyz
colnames(xyz)
rownames(xyz)
```

Finally, if the vectors aren't even of the same lengths we have

##Lists

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c("a", "a", "b")
z <- c(T, T)
xyz <- list(x, y, z)
xyz
```

lists are displayed quite differently from the other formats. Here are a number of commands:

```{r}
length(xyz)
names(xyz) <- c("Count", "Letter", "Married?")
xyz
```


