---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## Data Formats 

### Vectors

here are some useful commands for vectors:

```{r}
x <- c(1, 2, 3, 4, 5, 6)
x
length(x)
names(x) <- LETTERS[1:6]
x
```

If we have several vectors of the same type and length which belong together we can put them in a 

### Matrix

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c(4, 2, 3, 5, 3, 4)
z <- c(3, 4, 2, 3, 4, 2)
cbind(x, y, x)
rbind(x, y, z)
```

here are several other ways to make a matrix:

```{r}
matrix(x, 2, 3)
matrix(x, ncol=2)
matrix(0, 3, 3)
diag(3)
```

we can control how the matrix is filled in:

```{r}
matrix(1:8, nrow=2)
matrix(1:8, nrow=2, byrow = TRUE)
```

just like vectors a matrix is all of the same type, with R doing type conversion if necessary:

```{r}
matrix(c(1, 2, "A", 3), 2, 2)
```

useful commands for R matrices are:

```{r}
x <- matrix(c(1, 2, 3, 4, 5, 6), 2, 3)
x
dim(x)
nrow(x)
ncol(x)
dimnames(x) <- list(c("A", "B"), c("a", "b", "c"))
x
colnames(x)
rownames(x)
colnames(x) <- c("Height", "Age 1", "Age 2")
x
```

Actually I would recommend

```{r}
colnames(x) <- c("Height", "Age.1", "Age.2")
x
```

empty spaces in names are allowed but can on occasion lead to problems, so avoiding them is a good idea. Of course they are not nice as labels in graphs or as titles in tables, but it is usually easy to change those.

Sometimes a matrix is created by a function and given row and column names automatically, but we don't want it to have them. They can be removed with

```{r}
rownames(x) <- NULL
x
```

`r hl()$hr()`

Here is a strange way to make a matrix:

```{r}
x <- 1:6
x
dim(x) <- c(2, 3)
x
```

Why does this work? dim is an attribute of an object, by changing this attribute we are changing the object.

`r hl()$vspace()`

What happens if we try to make a matrix out of a vector that doesn't have the right number of entries?

```{r warning=TRUE}
matrix(1:5, 2, 2)
```

not surprisingly it just uses the elements needed.

```{r warning=TRUE}
matrix(1:3, 2, 2)
```

In this case R just starts all over again. This behavior is called *recycling* (in computer science, not just R)

In either case R gives a warning. Except in rare cases you should try to avoid these things, they are usually the consequence of bad programming.

### Arrays

an array is a k-dimensional matrix. For example

```{r}
array(1:8, dim=c(2, 2, 2))
```

Often these come about as a 3-way table: 

```{r}
x <- sample(1:3, size=20, replace=T)
y <- sample(c("F", "M"), size=20, replace = T)
z <- sample(c("a", "b", "c"), size=20, replace = T)
xyz <- table(x,y,z)
xyz
```

The commands for arrays are similar to those for matrices:

```{r}
dim(xyz)
dimnames(xyz)
```


### Data Frames

sometimes we have several vectors of the same length but of different types, then we can put them together as a data frame:

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c("a", "a", "b", "c", "a", "c")
z <- c(T, T, F, T, F, T)
xyz <- data.frame(x, y, z)
xyz
```

This type of data is very common in Statistics, and data frames have been the standard data type from its beginning. In general when you get data from a source like the internet it will be as a data frame.

`r hl()$vspace()`

The same commands as for matrices work for data frames as well:

```{r}
dim(xyz)
nrow(xyz)
ncol(xyz)
dimnames(xyz) <- list(letters[1:6], c("a", "b", "c"))
xyz
colnames(xyz)
rownames(xyz)
```

Say we want to add another column (variable) to the data frame:

```{r}
xyz[[4]] <- (1:6)^2
colnames(xyz)[4] <- "squares"
xyz
```

If we want to get rid of a column:

```{r}
xyz[[4]] <- NULL
```


There is a strange default behavior of data frames: they turn strings into factors:

```{r echo=FALSE}
options(stringsAsFactors = TRUE)
```

```{r}
df <- data.frame(x=1:5, y=letters[1:5])
str(df)
```

You can prevent this from happening, though:

```{r}
df <- data.frame(x = 1:5, 
                 y = letters[1:5],
                 stringsAsFactors = FALSE)
str(df)
```

In fact, this is what I want to happen almost all the time. So I change that option globally whenever I run R or RStudio. I will show you how to do that at some point.

If a data frame is all of the same type we can use *as.matrix* to turn it into a matrix.

**Exercise**

what does as.matrix() do when it is applied to a data frame with columns of different types?

`r hl()$hr()`

Finally, if the vectors aren't even of the same lengths we have

### Lists

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c("a", "a", "b")
z <- c(T, T)
xyz <- list(x, y, z)
xyz
```

lists are displayed quite differently from the other formats. Here are a number of commands:

```{r}
length(xyz)
names(xyz) <- c("Count", "Letter", "Married?")
xyz
```

Often we want to use a list inside a function to record various values. So we need to create an "empty" list of a certain length:

```{r}
x <- as.list(1:3)
x
```

and if we run out of space:

```{r}
x[[4]] <- "a"
x
```

Internally R stores all data as lists. In fact anything can be an element of a list:

#### **Example**

In Statistics we often have data of the form $(x, y)$ and we want to find a *linear model*. For example consider the data on *wine consumption and life expectancy*:

```{r}
kable.nice(wine)
attach(wine)
plot(Wine.Consumption, Heart.Disease.Deaths,
      xlab = "Wine Consumption",
      ylab = "Heart Disease per 100000")
```

**Note** The data set wine and the routine splot are part of a large collection of data sets and routines that I use in many of my courses. You can download the file at [http://academic.uprm.edu/wrolke/Resma3/Resma3.RData](http://academic.uprm.edu/wrolke/Resma3/Resma3.RData). 

Let's say we want to fit a linear model:

```{r}
fit <- lm(Heart.Disease.Deaths~Wine.Consumption)
plot(Wine.Consumption, Heart.Disease.Deaths,
      xlab = "Wine Consumption",
      ylab = "Heart Disease per 100000")
abline(fit)
```

Let's say we want to save the data and the fit for future use:

```{r}
wine.all <- list(data=wine, fit=fit)
length(wine.all)
```

data frames are lists, with the additional requirement that each column has the same length. But a "column" need not be what you think:

```{r}
z <- list(fit, fit, fit)
length(z)
df <- data.frame(x=1:3, y=letters[1:3])
df$z <- z
dim(df)
```

This works, but I don't recommend it. A list is likely a better option here.

**Example**

Say we have the following data set: in each of 10 experiments 5 measurements in five different locations coded as A-E were taken. The result was stored as a list, with each set of measurements an element:

```{r echo=FALSE}
results <- as.list(1:10)
for(i in 1:10) {
  results[[i]] <- round(rnorm(5, 100, 20), 1)
  names(results[[i]]) <- LETTERS[1:5]
}  
names(results) <- paste0("Experiment ", 1:10)
```

```{r}
results
```

Now we want to find the means by locations. One way would be to loop over the list elements, but much easier is:

```{r}
Reduce(`+`, results)/length(results)
```

