---
title: Answers to Exercises
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

These are the answers to the exercise questions:

#2.1 Data Types

What is the result of 

```{r eval=FALSE}
c(1, FALSE)
c("A", FALSE)
c(1L, FALSE)
-1 < FALSE
1 == "1"
```

```{r}
ltrs <-  c("a", "b", "b", "c", "c", "c")
f1 <- factor(ltrs)
f1
levels(f1) <- c("c", "b", "a")
f1
```

#2.2 Data Formats

what does as.matrix() do when it is applied to a data frame with columns of different types?

```{r}
df <- data.frame(x=1:3, y=letters[1:3])
as.matrix(df)
```

the usual type conversion

#2.3 Generate Vectors


What does this do?

```{r}
rep(1:10, 10:1)
```

What (if anything) does this do?

```{r}
(1:3 %o% 1:3) %o% 1:2
```

#2.4 Subsetting

What does this do?

```{r}
x <- 1:10
x[]
```

Nothing!


The data set *upr* (part of *Resma3.RData*) has the application information provided to the University of all students that were eventually accepted between 2003 and 2013. Here are the first three students:

```{r}
head(upr, 3)
```

How many female students applied in either 2010 or 2011, had a highschool GPA of at least 3.0 and a freshman GPA between 3.0 and 3.5?

```{r}
df <- upr[upr$Gender=="F", ]
df <- df[df$Year>=2010 & df$Year<=2011, ]
df <- df[df$Highschool.GPA>=3.0, ]
df <- df[df$Freshmen.GPA>=3.0 & df$Freshmen.GPA<=3.5, ]
df <- df[df$Class.Facultad!="INGE", ]
df[sample(1:nrow(df), size=10), 
   c("Gender", "Year", "Highschool.GPA",
     "Freshmen.GPA", "Class.Facultad")]
nrow(df)
```


why did I use 

logfac[c("2", 5, 30, 301, 30)]

and not

logfac[c(2, 5, 30, 301, 30)]

Vector indexes start at 1, not 0

#2.5 Random Numbers

How can we randomly select 20 rows of the **upr** data set?

```{r}
upr[sample(1:nrow(upr), size=20), 1:4]
```

How can we find Bootstrap samples of the **upr** data set?

```{r}
I <- sample(1:nrow(upr), 
            size = nrow(upr),
            replace = TRUE) 
head(upr[, 1:4])
```

Generate 10000 variates from a Binomial distribution with n=10, p=0.25 and compare the relative frequencies with the theoretical probabilities.

```{r}
x <- rbinom(1e4, 10, 0.25)
x.tbl <- round(as.numeric(table(x))/1e4, 3)
x.vals <- sort(unique(x))
x.prob <- round(dbinom(x.vals, 10, 0.25), 3)
data.frame(x=x.vals, 
           Simulation=x.tbl, 
           Theory=x.prob)  
```

generate 1000 variates from a Beta (0.5, 3) distribution, draw the histogram with 50 bins and add the density curve.

```{r}
x <- rbeta(1e3, 0.5, 3)
hist(x, 50, freq=FALSE)
f <- function(x) dbeta(x, 0.5, 3)
curve(f, 0, 1, add=TRUE, col="blue")
```

#2.6 Writing Functions, Programming

What does this function return?

```{r}
f <- function(x=y) {
  y <- 10
  x
}
f()
```

What is the difference between these two functions:

```{r}
f1 <- function(x) {
  if(x<10) return(0)
  x
}
f2 <- function(x) {
  ifelse(x<10, 0, x)
}
f1(1:20)
f2(1:20)
```

We already encountered one such infix function. What was it?

%o% outer product with *

Write a replacement function that adds n NA's to a vector

```{r}
`add.n<-` <- 
  function(x, value) {
    x <- c(x, rep(NA, value))
    x
  }
x <- 1:5
add.n(x) <- 3
x
```

#2.7 Object-Oriented Programming

#2.8 Vector Arithmetic, Linear Algebra

#2.9 Character Manipulation

The *Resma3* data set *worldpopulation* has some info on the population sizes of 227 countries. 

How many names of countries end with "o"? 

```{r}
co <- worldpopulation$Country
co[endsWith(co, "o")]
```

Let's say we have a vector x of numbers and we want to display them as a row in an html table. Write a single command that creates the needed html code

```{r}
x <- 1:5
paste0("<tr>", 
       paste0("<td>", x, "</td>", collapse = ""),
       "<tr>")
```

#2.10 Data Input/Output, Transfering R Objects

How many a's are there?

```{r, comment="", echo=FALSE}
x <- sample(letters[1:4], size=150, replace=T)
cat(x, "\n")
```

high-light letters, copy-paste, run

```{r eval=FALSE}
table(scan("clipboard", what="char"))
```

#2.11 Graphs

#2.12 Model Notation

for the **upr** data what is the coefficient of Highschool.GPA in a model that predicts Freshmen.GPA from Aptitud.Verbal, Aptitud.Matem, Aprov.Ingles and  Highschool.GPA with all second-order interactions except for Highschool.GPA-Aprov.Ingles?

```{r}
z <- summary(lm(Freshmen.GPA ~ (Aptitud.Verbal + Aptitud.Matem + Aprov.Ingles + Highschool.GPA)^2 - Highschool.GPA:Aprov.Ingles, data=upr))
z[[4]][5, 1]
```

#2.13 Numerical Methods

#2.14 Environments, Libraries

#2.15 Customizing R: .First, .Rprofile, Dropbox

#2.16 List of Important base R Commands

#3.1 Graphics with ggplot2

The data set *highways* has the rate of traffic accidents and other information for 39 highways.

Draw this graph:

```{r}
ggplot(aes(x=len, y=rate), data=highways) +
  geom_point()
```

Draw this graph:

```{r}
ggplot(aes(x=len, y=rate), data=highways) +
  geom_point() +
  labs(x = "Length of Highway", 
       y = "Rate of Accidents") +
  labs(title = "Rate decreases with length of highway") +
  geom_smooth(method="lm", se=FALSE)
```

#3.2 C++ with Rcpp

#3.3 Parallel and gpu Computing

#3.4 Input/Output Revisited

#3.5 The pipe, dplyr, tibbles, tidyverse
  
#3.6 Character Manipulation with strngr

#3.7 Dates with lubridate

#3.8 Factors with forcats

#3.9 Iteration with purrr

#3.10 Interactive Apps with shiny

#3.11 Version Control and Collaboration with github


#4.1  Estimation

#4.2  The Bootstrap

#4.3  Curve Fitting / Regression

#4.4  Classification

#4.5  Bayesian Analysis

