---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`
`r hl()$basefontsize()`

## Generating Objects

### Vectors

There are numerous ways to generate vectors which have some structure. Here are some useful commands:

The easiest way to do this is to use the *c* (concatenate) command:

```{r}
x <- c(0, 2, 3, 1, 5)
x
```

to make regular sequences use ":" (read "to")

```{r}
1:5
-2:2
```

and then we can combine these:

```{r}
c(1:5, 11:15)
```

`r hl()$vspace()`

there are also a number of commands for this purpose:

-  seq

```{r}
seq(1, 10, 1)
seq(1, 10, 1/2)
seq(0, 10, length=20)
```

-  sequence

this creates a series of sequences of integers each ending by the numbers given as arguments:

```{r}
sequence(10)
sequence(c(2, 5, 3))
```

-  rep

```{r}
rep(1, 10)
rep(1:3, 10)
rep(1:3, each=3)
rep(c("A", "B", "C"), c(4, 7, 3))
rep(1:3, rep(5, 3))
```

**Exercise**

What does this do?

```{r eval=FALSE}
rep(1:10, 10:1)
```

-  gl

The function *gl* (generate levels) is very useful because it generates regular series of factors. The usage of this function is gl(k, n) where k is the number of levels (or classes), and n is the number of replications in each level. 

Two options may be used: length to specify the number of data produced, and labels to specify the names of the levels of the factor. 

```{r}
gl(3, 5)
gl(3, 5, length=30)
gl(2, 6, label=c("Male", "Female"))
```


-  expand.grid

this takes a couple of vectors and writes them as a matrix with each combination.

```{r}
expand.grid(1:2, 1:3)
expand.grid(First=1:2, Second=1:3, Third=c("A", "B"))
```

there are a number of R routines who need the data in this format as arguments, so this is an easy way to convert them.

-  outer

This calculates the outer product of two vectors. It creates a matrix of length(x) x length(y) where $z_{ij} = f(x_i, y_j)$ for some function $f$

```{r}
x <- 1:3
y <- 1:5
outer(x, y, "*")
```

You can use any function you like:

```{r}
outer(x, y, function(x,y) {x^2/y})
```

If it is actual multiplication you want there is also a short hand:

```{r}
x %o% y
```

Want to practice your multiplication tables?

```{r}
x <- 1:10
names(x) <- x
x %o% x
```

**Exercise**

What (if anything) does this do?

```{r eval=FALSE}
(1:3 %o% 1:3) %o% 1:2
```

### Specialty Data

Some common objects are easy to create:

```{r}
letters
LETTERS
```


### Expressions

up to now we discussed how to generate data objects. Soon we will be talking about how to write your own functions. There is however also a type of object somewhat in between, so called *expressions*.

An expression is a series of characters which make sense for R. All valid commands are expressions. When a command is typed directly on the keyboard, it is then evaluated by R and executed if it is valid. In many circumstances, it is useful to construct an expression without evaluating it: this is what the function expression is made for. It is, of course, possible to evaluate the expression subsequently with eval().

```{r}
x <- 3; y <- 2.5; z <- 1 
exp1 <- expression(x / (y + exp(z))) 
exp1 
eval(exp1) 
```

Expressions can be used for many things. Here are two examples:

1.  I want to draw the graph of a function, including the equation:

```{r}
curve(x^2*exp(-x^2), -2, 2,
      main = expression(x^2*exp(-x^2)), 
      ylab = "")
```

2.  Do symbolic math. For derivatives we have the function *D*, which returns partial derivatives:

```{r}
D(exp1, "x")
eval(D(exp1, "x"))
D(exp1, "y")
eval(D(exp1, "y"))
D(exp1, "z")
eval(D(exp1, "z"))
```

**Exercise**

Use the D function to find $\frac{d^2}{dx^2} \frac{x^2}{1+x^3}|_{x=0}$

```{r echo=FALSE, eval=FALSE}
x <- 0
sexp <- expression(x^2/(1+x^3))
eval(D(D(sexp, "x"), "x"))
```


In general R's use as a symbolic language is very limited. There are of course many purpose built languages for this, such as Maple or Mathematica.


