---
title: Graphs
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center",
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

In this section we will discuss some graphs that are part of base R. Later we will study more advanced graphics, but it is still a good idea to know how to draw a base graph.

#Barcharts

Say we have the data on the number of freshman, sophomores etc at some college:

```{r, echo=FALSE}
student.levels <- 
  c("Freshman", "Sophomore", "Junior", "Senior")
students <- sample(student.levels, 
                   size=1000, 
                   replace = TRUE, 
                   prob = 6:3)
st.y <- factor(students, 
               levels = student.levels, 
               ordered = TRUE)
students.ord <- table(st.y)
kable(students.ord, col.names = c("Class", "Counts"))
```

```{r}
head(students)
```

So here we have a *categorical* variable. One popular graph for this type of data is the *pie chart*, 

BUT repeat after me: [pie charts are evil!!](http://www.storytellingwithdata.com/blog/2011/07/death-to-pie-charts)

A MUCH better alternative are barcharts:

```{r}
barplot(table(students))
```

**Note** that the argument to barplot has to be a table.

But this is not quite right, Sophomores should come second. In general, when we have a categorical variable which has an ordering we should turn it into a factor:

```{r}
students.fac <- 
  factor(students, 
         levels = c("Freshman", "Sophomore", 
                 "Junior", "Senior"), 
         ordered = TRUE)
tbl.students.fac <- table(students.fac)
barplot(tbl.students.fac)
```

There are a number of arguments we find in most graphics routines:

```{r}
barplot(tbl.students.fac, 
        main = "Our Students",
        xlab = "Class", 
        ylab = "Counts")
```

and then there are arguments specific to the graph:


```{r}
barplot(tbl.students.fac, 
        main = "Our Students",
        xlab = "Counts", 
        ylab = "", 
        horiz = TRUE, 
        density = 10,
        col = c("lightblue", "mistyrose",
                "lightcyan", "lavender"))
```

Notice that we can't see all the y labels. It would be better to have them horizontal as well. But we also need to make space for them:

```{r}
par(mai=c(1,2,1,1))
barplot(tbl.students.fac, 
        main = "Our Students",
        xlab = "Counts", 
        ylab = "", 
        horiz = TRUE, 
        density = 10,
        las = 1,
        col = c("lightblue", "mistyrose",
                "lightcyan", "lavender"))
```

Here I used the *par* command to set some parameters for how the graph is drawn. There is a very long list of such parameters see ?par

#Histogram

```{r}
x <- rnorm(1000, 100, 20)
hist(x, 
     breaks = 50, 
     main = "IQ",
     xlab = "IQ", 
     ylab = "Counts")
```

Often the hist command is useful for its side effect of counting:

```{r error=TRUE}
range(x)
hist(x, 
  breaks = c(0, 50, 80, 100, 120, 150, 250), 
  plot = FALSE)$counts
```

Often we want to compare a histogram with a theoretical curve, say a probability density. Then we can use the curve command with the argument add=TRUE:

```{r}
x <- c(rnorm(1500, 100, 20), rnorm(1000, 140, 10))
hist(x, 
     breaks = 50, 
     main = "", 
     freq=FALSE)
f <- function(x) 
   1.5*dnorm(x, 100, 20) + dnorm(x, 140, 10)
I <- integrate(f, 0, 200)$value
curve(f(x)/I, from=min(x), to=max(x), 
      add = TRUE, 
      col = "blue", 
      lwd = 2)
```

#Empirical Distribution Function

```{r}
plot(ecdf(x))
```


#Boxplot

```{r}
boxplot(x, horizontal = TRUE)
attach(mothers)
boxplot(Length~Status)
```

#Normal Probability Plot

```{r}
qqnorm(x)
qqline(x)
```

#Scatterplot


```{r echo=FALSE}
rm(plot)
```


```{r}
attach(faithful)
plot(Eruptions, Waiting.Time, 
     xlab = "Length of last eruption",
     ylab = "Time until next eruption", 
     main = "Old Faithful Guyser")
```

Let's add the least squares regression fit to the graph:

```{r}
plot(Eruptions, Waiting.Time, 
     xlab = "Length of last eruption",
     ylab = "Time until next eruption", 
     main = "Old Faithful Guyser")
abline(lm(Waiting.Time~Eruptions), lwd=2, lty=2)
```

There are a lot of different plotting symbols:

```{r}
plot(0:16, rep(0, 17), ylim=c(0,1), 
     type="n", xlab="", ylab="", axes=F)
for(i in 1:15) {
  for(k in 1:3) {
    points(i, c(0.8, 0.5, 0.2)[k], pch=i+(k-1)*15)
    text(i, c(0.9, 0.6, 0.3)[k], label=i+(k-1)*15, adj = 0) 
  }
}  
```

Notice the setup of an empty graph!

Also, different line types:

```{r}
plot(0:7, rep(0, 8), ylim=c(0,1), 
     type="n", xlab="", ylab="", axes=F)
for(i in 1:6) {
  segments(i, 0.2, i, 1, lty=i)
  text(i, 0.1, i)
}

```

Consider the data set mtcars, which is part of base R. It has data on 32 different cars:

```{r}
head(mtcars)
```

say we wish to study the relationship of mpg (Miles per Gallon) and hp (Horsepower) but also include information on the number of cylinders (either 4, 6 or 8):

```{r message=FALSE}
attach(mtcars)
cols <- rep("blue", 32)
cols[cyl==6] <- "green"
cols[cyl==8] <- "red"
plot(hp, mpg, 
     xlab = "Horsepower", 
     ylab = "Miles per Gallon",
     col = cols, pch = 20)
legend(250, 33, paste(c(4, 6, 8), "Cylinders"),
       pch = 20, 
       col = c("blue", "green", "red"))
```

Let's add the least squares regression lines for each cylinder separately. We could use the *abline(fit)* command, but notice that the hp ranges for the three cylinder types are quite different. Using abline would draw the fit all the way across the graph. Here is a better solution: 

```{r message=FALSE}
cls <- c("blue", "green", "red")
plot(hp, mpg, 
     xlab = "Horsepower", 
     ylab = "Miles per Gallon",
     col = cols, pch = 20)
legend(250, 33, paste(c(4, 6, 8), "Cylinders"),
       pch = 20, 
       col = cls)
for(i in c(4, 6, 8)) {
  fit <- lm(mpg[cyl==i]~hp[cyl==i])
  x <- range(hp[cyl==i])
  y <- coef(fit)[1] + coef(fit)[2]*x
  segments(x[1], y[1], x[2], y[2], col=cls[(i-2)/2])
}
```


#Multiple Graphs

sometimes we want to combine several graphs into one

```{r}
par(mfrow=c(2, 2))
hist(rnorm(1000), 50, 
     main="Normal Distribution",
     ylab="", xlab="")
hist(rt(1000, 2), 50, main="t Distribution, df=2",
     ylab="", xlab="")
hist(runif(1000), 50, main="Uniform Distribution",
     ylab="", xlab="")
hist(rchisq(1000, 2), 50, 
     main = expression(paste(chi^2, 
            " Distribution, df=2")),
     ylab="", xlab="")
```

This works fine for rectangular arrays. For more complicated graphs we have the *layout* command. Let's create a scatterplot of mpg by hp with marginal boxplots:

```{r}
layout(matrix(c(2, 0, 1, 3), 2, 2, byrow = TRUE), 
             c(3, 1), c(1, 3), TRUE)
par(mar = c(3, 3, 1, 1))
plot(hp, mpg)
par(mar = c(0, 3, 1, 1))
boxplot(hp, axes = FALSE, horizontal = TRUE)
par(mar = c(3, 0, 1, 1))
boxplot(mpg, axes = FALSE)
```

#Functions that create graphs

When I write a paper or a talk or anything else that requires some graphs I always write a routine *plot.mytalk*, which has everything to recreate every graph. This is a great way to not only make changes if necessary but also to remind myself what I did back years ago when I wrote it.

So I might have the following function to do the graph above:


```{r eval=FALSE}
figure1 <- function() {
  layout(matrix(c(2, 0, 1, 3), 2, 2, 
                byrow = TRUE), 
         c(3, 1), c(1, 3), TRUE)
  par(mar = c(3, 3, 1, 1))
  plot(hp, mpg)
  par(mar = c(0, 3, 1, 1))
  boxplot(hp, axes = FALSE, horizontal = TRUE)
  par(mar = c(3, 0, 1, 1))
  boxplot(mpg, axes = FALSE)
  
}  
```


#Graphs that don't look like Graphs

we can use R to make other types of pictures. Consider this diagram, which I use to illustrate the topic of Cause vs Effect:

```{r, echo=FALSE}
draw.cause.effect <- function() {
  plot(c(10, 100), c(40, 100), 
       axes = FALSE, 
       xlab = "", ylab = "", 
       type = "n")
  text(15, 90, "Cause-Effect", 
       cex=1.2, adj = 0)
  text(c(20, 40), c(65, 65), c("X", "Y"), 
       cex=1.1, adj=0)
  arrows(25, 65, 38, 65)
  text(56, 90, "Confounding Variable", 
       cex=1.2, adj=0)
  text(c(60, 70, 80), c(50, 75, 50), c("X", "Z", "Y"),
       cex=1.1, adj=0)
  arrows(70, 70, 62, 55)  
  arrows(70, 70, 78, 55)  
  
}
draw.cause.effect()
```

it is actually done with these commands:

```{r eval=FALSE}
plot(c(10, 100), c(40, 100), axes=F, 
     xlab="", ylab="", type="n")
text(15, 90, "Cause-Effect", 
     cex=1.2, adj = 0)
text(c(20, 40), c(65, 65), c("X", "Y"), 
     cex=1.1, adj=0)
arrows(25, 65, 38, 65)
text(56, 90, "Confounding Variable", 
     cex=1.2, adj=0)
text(c(60, 70, 80), c(50, 75, 50), c("X", "Z", "Y"),
     cex=1.1, adj=0)
arrows(70, 70, 62, 55)  
arrows(70, 70, 78, 55)  
```

An important example of this are maps:

```{r}
library(maps)
map("usa")
text(-87.623177, 41.881832, "Chicago", cex=1.3)
```

People have done some very funny things with this. Run this code in the console (you need to install the library *cowsay* first)

```{r eval=FALSE}
library(cowsay)
say("Hello world!", by = "cow")
```



#Printing Graphs

Often we need to save a graph as a png or a postscript file, so we can include it in a webpage or a latex document. I have to do this often enough I wrote a routine for it:

```{r}
graph.out <-
function (f, foldername, graphname, format = "png") 
{
  file <- paste0(foldername, graphname, ".", format)
  cat(file)
  if (format == "png") 
      png(file)
  if (format == "pdf") 
      pdf(file)
  if (format == "ps") 
      postscript(file, horizontal = F, pointsize = 17)
  if (format == "eps") {
      setEPS()
      postscript(file, horizontal = F, pointsize = 17)
  }
  f()
  dev.off()
  
}
```

here *f* is a function that produces a graph, so it might be called like this:

```{r eval=FALSE}
f <- function(x) hist(rnorm(1000), 50)
graph.out(f, "C:/mygraphs", "myhist")
```

and now there should be a file myhist.png in the folder C:/mygraphs.

 
 

