---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/Resma3/Resma3.RData"))

require(knitr)
require(ggplot2)
require(grid)
require(leaps)
```
`r rmd$basefontsize()`

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

#History

##S

S is a statistical programming language developed primarily by John Chambers and (in earlier versions) Rick Becker and Allan Wilks of Bell Laboratories. The aim of the language, as expressed by John Chambers, is "to turn ideas into software, quickly and faithfully".

The first working version of S was built in 1976, and operated on the GCOS operating system. At this time, S was unnamed, and suggestions included Interactive SCS (ISCS), Statistical Computing System, and Statistical Analysis System (which was already taken: see SAS System). The name 'S' (used with single quotation marks, until 1979) was chosen, as it has the common letter used in statistical computing, and is consistent with other programming languages designed from the same institution at the time (namely the C programming language).

By 1988, many changes were made to S and the syntax of the language. The New S Language (1988 Blue Book) was published to introduce the new features, such as the transition from macros to functions and how functions can be passed to other functions (such as apply).

In the late 1990s At&T sold the S license to a private company that started to charge a lot of money for it. This lead to

##R

R is an implementation of the S programming language combined with lexical scoping semantics inspired by Scheme. There are some important differences between S andR, but much of the code written for S runs unaltered.

R was created by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand, and is currently developed by the R Development Core Team, of which Chambers is a member. R is named partly after the first names of the first two R authors and partly as a play on the name of S. The project was conceived in 1992, with an initial version released in 1995 and a stable beta version in 2000.

##RStudio

RStudio is a free and open-source integrated development environment (IDE) for R, a programming language for statistical computing and graphics. RStudio was founded by JJ Allaire, creator of the programming language ColdFusion. Hadley Wickham is the Chief Scientist at RStudio.

Work on RStudio started at around December 2010,[9] and the first public beta version (v0.92) was officially announced in February 2011.[1] Version 1.0 was released on 1 November 2016. Version 1.1 was released on 9 October 2017.

R can be run by itself using the command line interface, but it is usually much nicer to use RStudio as a front end.

##Getting Started

Once R (or Rstudio) is opened you will see the >. This is the *command prompt*. It means R is waiting for you to do something!

Don't like >? No sweat, you can change it (although I don't recoomend changing the prompt). Here is how:

```{r eval=FALSE}
options()(prompt="@")
```

In fact, you can change almost anything with the options command.

```{r}
length(options())
```

shows there are 63 of them. Two you might want to remember are

```{r}
options()["digits"]
pi
options(digits=15)
pi
```

and

```{r}
options()$warn
```

this tells R to warn about certain things. Those are not errors, just warnings, and usually they are ok. Sometimes though they are not needed but can be a nuisance, and then you can turn them off with

```{r}
options(warn=-1)
```


##Assignments

The basic assignment character in R is <-

```{r}
x <- 1
x
```

*Note* = also works:

```{r}
x = 2
x
```

but for some good reasons is not the best way to go.

##Help!

R has a help command. Let's say we want to find out about the *mean* command:

```{r, eval=FALSE}
help(mean)
```

or even quicker

```{r eval=FALSE}
?mean
```

Another good command is *arg*, which lists the arguments of a routine:

```{r}
args(splot)
```

*splot* is not a command that is part of R but one I wrote myself to make nice looking scatterplot. Unfortunately args often doesn't work well for basic R commands:

```{r}
args(mean)
```

##ls, rm

use the ls command to get a listing of the current objects in the worksheet:

```{r}
ls()[1:5]
```

If there are many objects (the worksheet I have open right now has over 250!) you can specify part of the name:

```{r}
ls(pattern = "^sa")
```

the ^ in front means that ls will list all object whose name starts with sa. This is what is called a *regular expression*. These are general (non R) rules for specifying patters. They are a science all their own! 

Often you create an object that is only need for a short time. If you want to get rid of it use rm (remove)

```{r}
x <- 1:10
sum(x^3)
rm(x)
```

you can also remove many objects in one step. Let's say we have worked on a problem for a while and we created a number of objects. All of the start with "my". Now we can do this

```{r eval=FALSE}
a <- ls(pattern = "^my")
rm(list=a)
```

In many of my projects I have a routine called *cleanup*:

```{r}
cleanup <- function () 
{
  oldstuff <- c("f","f1","n","m","a","A","b", "B", "i","j", "plt","plt1","plt2","x","xx","y","z", "xy")
  rm(list=oldstuff, envir = .GlobalEnv)
}
```

Notice the argument envir. We will talk about what that means soon.
