---
title: Homework 3
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
library(wolfr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

In this homework you will need to use data from *Resma3.RData*. Go to our homepage and download it. Save it in "folder". Include the following chunk in your Rmd:

```{r eval=FALSE}
load("c:/folder/Resma3.RData")
```

**Problem 1**:

The data set *upr* (part of *Resma3.RData*) has the application information provided to the University of all students that were eventually accepted between 2003 and 2013. 

a.  How many students applied to Arts & Sciences?

b.  How many male students applied to Arts & Sciences?

c.  How many students applied to Arts & Sciences in the years 2008-2010 (inclusive)

d.  How many female students applied in either 2010 or 2011, had a highschool GPA of at least 3.0 and a freshman GPA between 3.0 and 3.5?

**Problem 2**

Generate the following sequences. Do not use c() alone. There is more than one solution to each, you just need to find one.

a. 

```{r echo=FALSE}
rep(1:3, c(2, 3, 1))
```

b. 

```{r echo=FALSE}
seq(0, 1, 0.2)
```

c.

```{r echo=FALSE}
sequence(3:5)
```

d.

```{r echo=FALSE}
rep(letters[1:4], rep(2, 4))
```

e. 

```{r echo=FALSE}
rep(LETTERS[5:8], c(4, 2, 2, 4))
```

**Problem 3**

In class we used the lm command to fit a linear model. The result of a call to lm is a rather complicated object with a lot of information. Let's study it in some detail:

```{r}
fit <- lm(Heart.Disease.Deaths~Wine.Consumption,
          data=wine)
summary(fit)
```

a. What data format does the fit object have?

b. How many elements does it have and what are their names?

c. How can you get the coefficients (`r coef(fit)[1]` and `r coef(fit)[2]`) alone? (Maybe so thay can be used in an inline code like I just did!)

d. In the table above one of the items is the t value of Wine.Consumption (18.833). How can you get that value alone?
