---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
runtime: shiny
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Installation and Updating

You can get a free version of R for your computer from a number of sources. The download is about 70MB and setup is fully automatic. Versions for several operating systems can be found on the R web site

[https://cran.r-project.org](https://cran.r-project.org)


*Note* 

-  the one item you should change from the defaults is to install R into a folder under the root, aka C:\\R

-  You might be asked at several times whether you want to do something (allow access, run a program, save a library, ...), always just say yes!

-  You will need to connect to a reasonably fast internet for these steps.

-  This will take a few minutes, just wait until the > sign appears.

`r hl()$hr()`

**FOR MAC OS USERS ONLY**

There are a few things that are different from MacOS and Windows. Here is one thing you should do:

Download XQuartz - XQuartz-2.7.11.dmg  
Open XQuartz      
Type the letter R (to make XQuartz run R)        
Hit enter
          Open R 
          Run the command .First()          
Then, every command should work correctly. 


###RStudio

We will run R using an interface called **RStudio**. You can download it at [RStudio](https://www.rstudio.com/). 


###Updating

R releases new versions about every three months or so. In general it is not necessary to get the latest version every time. Every now and then a package won't run under the old version, and then it is time to do so. In essence this just means to install the latest version of R from CRAN. More important is to now also update ALL your packages to the latest versions. This is done simply by running

```{r, eval=FALSE}
update.packages(ask=FALSE, dependencies=TRUE)
```



