---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## The pipe, dplyr, tibbles, tidyverse

### The pipe

The traditional workflow of R comes in large part from other computer languages. So a typical sequence would be like this:

```{r}
df <- mtcars
df1 <- subset(df, hp>70)
ggplot(data=df1, aes(disp, mpg)) + 
  geom_point()
```

This is not how we think, though. That would go something like this:

-  take the mtcars data set  
-  then pick only cars with hp over 70  
-  then do the scatterplot of mpg vs. disp

In addition there is also the issue that we had to create an intermediate data set (df1).

The *pipe* was invented to fix all of these problems.

The basic package to use piping is

```{r}
library(magrittr)
```

invented by Stefan Milton. The basic operator is %>%. The same as above can be done with

```{r}
mtcars %>%
  subset(x=., hp>70) %>% #R knows what hp is!
  ggplot(data=., aes(disp, mpg)) + 
     geom_point()
```

In principle the pipe can always be used in this way:

```{r}
x <- rnorm(10)
mean(x)
x %>% 
  mean()
```

Notice that here we called both mean and round without a needed argument. In principle the pipe will always use the data on the left of %>% as the first argument of the command on the right.

**Exercise**

Consider the following operation:

```{r}
x <- c(0.109, 0.359, 0.63, 0.996, 0.515, 0.142, 0.017, 0.829, 0.907)
round(exp(diff(log(x))), 1)
```

write the same using the pipe

```{r echo=FALSE, eval=FALSE}
x %>% log() %>%
    diff() %>%
    exp() %>%
    round(1)
```

`r hl()$hr()`

At first it may not seem like writing x %>% f() is any easier than writing f(x), but this style of coding becomes very useful when applying multiple functions; in this case piping will allow one to think from left to right in the logical order of functions rather than from inside to outside in an ugly large nested statement of functions.

`r hl()$hr()`

The pipe is only a few years old, but there are already many packages that take advantage of it. The most important one, and a package useful in and of itself, is 

```{r}
library(dplyr)
```

written by Hadley Wickham. In essence it is a replacement for the *apply* family of R routines. We can also write the above with

```{r}
mtcars %>%
  filter(hp>70) %>%
  ggplot(aes(disp, mpg)) + 
     geom_point()
```

Notice how *filter* is aware of the pipe, it doesn't need to be told that it is supposed to work with mtcars. So far, *ggplot* is not fully pipe aware (otherwise we could have written %>% geom_point()), but this will change in the near future.

### tibbles

dataframes have been the main data format of R since its beginnings, and are likely to stay that way for a long time. They do, however have some shortcomings. Among other things, when you type the name of a dataframe and hit enter, all of it is shown, even if the data set is huge. On the other hand, interesting information such as the data types of the columns is not shown. To help with these (and some other) issues the data format *tibble* was invented. We can turn a dataframe into a tibble with

```{r}
tmtcars <- as.tbl(mtcars)
tmtcars
```

so we have all relevant information about the data set: its size (32x11), the variables and their formats, and the beginning of the data set.

*tibbles* are also designed to work well with piping and with the package *dplyr*.

If you want to create a tibble from scratch use:

```{r}
tibble(x=1:5, y=x^2)
```

Also, tibbles never use row.names, and it only recycles vectors of length 1. This is because recycling vectors of greater lengths is a frequent source of bugs.

### *dplyr* library

We have already seen the filter command, the *dplyr* version of subset. Here are the most important *dplyr* commands: 

-  filter  selects part of a data set by conditions (base R command: subset)
-  select 	selects columns  (base R command: [ ])
-  arrange 	re-orders or arranges rows  (base R commands: sort, order)
-  mutate 	creates new columns  (base R commands: any math function)
-  summarise 	summarises values  (base R commands: mean, median etc)
-  group_by 	allows for group operations in the  “split-apply-combine” concept (base R command: none)


#### **Example**: babynames

The library *babynames* (also by Hadley Wickham) has the number of children of each sex given each name for each year from 1880 to 2015 according to the US census. All names with more than 5 uses are given.

We want to do the following:

-  take the names  
-  pick out all of those that start with "W"  
-  separate the genders  
-  find the total for each year  
-  do the line graph

```{r}
library(babynames)
babynames %>%
    filter(name %>% substr(1, 1) %>% equals("W")) %>%
    group_by(year, sex) %>%
    summarise(total = sum(n)) %>%
    ggplot(data = ., aes(year, total, color = sex)) +
      geom_line() + 
      labs(color="Gender") +
      ggtitle('Names starting with W') 
```

How often is my name used for a baby in the US?

```{r}
babynames %>%
    filter(name == "Wolfgang") %>%
    ggplot(data = ., aes(year, n)) +
      geom_line()
```

Looks like my name is getting more popular (even if it is still rare)!



What were the most popular girls names each year?

```{r}
babynames %>%              # take babynames
    filter(sex=="F") %>%   # then pick girls only
    group_by(year) %>%     # then separate the years
    mutate(M=max(n)) %>%   # then find most often used name
    filter(n==M) %>%       # then pick only those rows
    ungroup() %>%          # then join data back together
    select(name) %>%       # then select names only
    table() %>%            # then count how often each happened  
    sort(decreasing = TRUE) %>% # then organize data
    cbind()               #  then turn data around for easier reading  
```

Let's say we want to save a data set made with the pipe. Logically we should be able to do this

```{r error=TRUE}
babynames %>%              # take babynames
  filter(name=="Wolfgang") %>%   # then pick me
  wolfgangs                # then give new data set a name
```

but that results in an error, only functions can be used in a pipe. So it is done like this:
 
```{r}
wolfgangs <- babynames %>%              # take babynames
  filter(name=="Wolfgang")    # then pick me
print(wolfgangs, n=3)
```

This unfortunately breaks the logic of piping. There is a better way, though. Just remember the logic of the assignment character <-, it's an arrow!


```{r}
babynames %>%              # take babynames
  filter(name=="Wolfgang") ->   # then pick me
  wolfgangs               # then assign it a name
print(wolfgangs, n=3)
```

`r hl()$hr()`

Here is a common problem: say you have these two data sets:

```{r echo=3:4}
students1 <- tibble(name=c("Alex", "Ann", "Marie"), 
                    exam1=c(78, 85, 93))
students2 <- tibble(name=c("Alex", "Ann", "Marie"), 
                    exam2=c(75, 89, 97))
students1
students2
```

and we want to join them into one data set:

```{r}
students1 %>%
  left_join(students2)
```

`r hl()$hr()`

Let's say we want to find the times out of 100000 that the most popular names occurred in the 2015:

```{r}
babynames %>%
  filter(year==2015) %>%
  mutate(freq=round(n/sum(n)*100000)) %>%
  select(name, freq) %>%
  arrange(desc(freq)) %>%
  print(n=5)
```

so the *mutate* command let's us calculate new variables and the *arrange* command let's us change the order of the rows.

### The tidyverse

*ggplot2* and *dplyr* are two of a number of packages that together form the *tidyverse*. They are centered around what is called *tidy data*. For a detailed discussion go to [https://www.tidyverse.org/](https://www.tidyverse.org/).

The core packages are

-  ggplot2  
-  dplyr  
-  tidyr  
-  readr  
-  purrr  
-  tibble  
-  stringr  
-  forcats

but you can get all of them in one step with

```{r eval=FALSE}
install.packages("tidyverse")
```

*tidy data* is defined as data were

1. Each variable you measure should be in one column.  
2.  Each different observation of that variable should be in a different row.  
3.  There should be one table for each "kind" of variable.  
4.  If you have multiple tables, they should include a column in the table that allows them to be linked.

This is essentially the definition of a data frame, but it is enforced even more so by the *tibbles* format. The theory behind tidy data was described by Hadley Wickham in the article [Tidy Data](https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf),  Journal of Statistical Software. The packages in the tidyverse are all written to have a consistent look and feel and work naturally with tidy data.

One big difference between dataframes and tibbles is that tibbles automatically ignore row names:

```{r}
head(mtcars, 3)
tbl.mtcars <- as.tbl(mtcars)
print(tbl.mtcars, n=3)
```

This of course is no good here, the names of the cars are important. One way to fix this is to use the *rownames_to_column* routine in the *tibbles* package:

```{r}
library("tibble")
mtcars %>%
   as.tbl() %>%
   rownames_to_column() %>%
   print(n=3)
```

One difficulty is to remember which routine is in what package. The best way is to simply load them all with

```{r eval=FALSE}
library(tidyverse)
```



