---
title: Iteration with purrr
                 \usepackage{float}
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

We have previously learned how to write loops and how to use apply. In this section we will learn about another variation on that theme.

We will need

```{r}
library(tidyverse)
```

Say we have the following problem. We have a dataframe

```{r}
df <- tibble(
  a1 = round(rnorm(100), 1)
)
for(i in 2:50) {
  df[[i]] <- round(rnorm(100, i), 1)
  colnames(df)[i] <- paste0("a", i)
}
head(df, 2)
```

and we want to find the mean of the 50 columns.

We can of course use a loop:

```{r}
df.mean <- rep(50, 0)
for(i in seq_along(df))
  df.mean[i] <- mean(df[[i]])
df.mean[1:5]
```

or we can use apply:

```{r}
apply(df, 2, mean)[1:5]
```

and then there is the *purrr* routine

```{r}
map_dbl(df, mean)[1:5]
```

What is the advantage of map over apply or even just the loop? One is that map is designed to work with the pipe:

```{r}
df %>% 
  map_dbl(mean) ->
  out
out[1:5]
```

Another is that *map* has a number of nice shortcuts.

On a number of occasions we already faced the following problem: we have a data set with two quantitative variables and a categorical one, and we want to fit separate linear models to each value of the categorical one.

If the goal is a scatterplot this is easy:

```{r}
ggplot(data=mtcars, 
       aes(wt, mpg, color=factor(cyl))) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(x="Weight", y="MPG", color="Cylinders")
```

but what if we want the actual models? Here is an easy way to do this:

```{r}
mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg~wt, data=.)) ->
  mtcars.models
summary(mtcars.models[[1]])
```

Notice the shorthand *\~lm(mpg\~wt)* instead of the long form *function(df) lm(mpg\~wt, data=df)*

Now say we want to see the R^2's:

```{r}
mtcars.models %>% 
  map(summary) %>% 
  map_dbl("r.squared") 
```

**Example:**

The data in the *gapminder* package has information on most of the countries in the world, including Puerto Rico. For a detailed explanation see its homepage at [https://www.gapminder.org/](https://www.gapminder.org/).

```{r}
library(gapminder)
```

Let's consider for a moment the life expectancy per country over time:

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
    geom_point() +
    geom_smooth(se=FALSE) +
    labs(x="Year", y="Life Expectancy")
```

Not surprisingly, the life expectancy has increase over the last 70 years or so. Overall we have

```{r}
fit <- lm(gapminder$lifeExp ~ gapminder$year)
summary(fit)
```

so the life expectancy increased an average of 0.33 years or 4 months per year from 1950 to 2010.

It does seem that the linear model is not quite right, and in fact

```{r}
ggplot(data.frame(x=fitted.values(fit), y=residuals(fit)), aes(x,y)) +
  geom_point() +
  labs(x="Fitted Values", y="Residuals")
```

but instead of pursuing this issue let's find a linear model for each country:

```{r}
gapminder %>% 
  group_by(country) %>% 
  nest() ->
  gap.country
gap.country
```

This splits the data set into 142 data sets, one for each country, but still organizes the whole as a tibble (rather than say as a list).

What is this column "data"? As it says, it is a tibble again. So we have a tibble with a colunm, each entry of which is again a tibble!

Let's check Puerto Rico:

```{r}
I.PR <- c(1:dim(gap.country)[1])[gap.country$country=="Puerto Rico"]
gap.country$data[[I.PR]]
```

which shows us the year by changes in Puerto Rico.

Next we find the least squares regression for each country:

```{r}
gap.country %>% 
  mutate(model = map(data, function(df) lm(lifeExp ~ year, data=df))) ->
  gap.country
gap.country
summary(gap.country$model[[I.PR]])
```

Now that we have the linear models we can use them to do the graph. The best way to do this is to undo the nest() command and get back to a standard dataframe:

First we add the fitted values and the residuals to the tibble. For this we need a new package called *modelr*

```{r}
library(modelr)
gap.country %>% 
  mutate(fits = map2(data, model, add_predictions)) %>% 
  mutate(resids = map2(data, model, add_residuals)) ->
  gap.country
gap.country
```

```{r}
fits <- unnest(gap.country, fits)
fits
resids <- unnest(gap.country, resids)
```

```{r}
fits %>% 
  ggplot(aes(year, pred)) +
    geom_line(aes(group=country), alpha=1/3) +
    geom_smooth(se=FALSE)
```

and we see that while the general trend is up there are actually exceptions!

Here are the residuals:

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group=country), alpha=1/3) +
    geom_smooth(se=FALSE)
```

and again we see that for some country the linear model fails badly.

(Notice the use of group instead of color. Any idea why?)

Here is an interesting view:

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group=country), alpha=1/3) +
    geom_smooth(se=FALSE) +
    facet_wrap(~continent)
```

For comparison let's redo this analysis, this time using base R only.

```{r}
df <- gapminder
n.countries <- dim(df)[1]
df$fitted <- rep(0, n.countries)
df$residuals <- rep(0, n.countries)
for(i in 1:n.countries) {
  z <- df[df$country==df$country[i], ]
  fit <- lm(z$lifeExp~z$year)
  df[df$country==df$country[i], "fitted"] <- fitted.values(fit)
  df[df$country==df$country[i], "residuals"] <- residuals(fit)
}
plot(df$year, df$lifeExp, type="n", xlab="Year", ylab="Fits")
for(i in 1:n.countries) {
  z <- df[df$country==df$country[i], ]
  lines(z$year, z$fitted)
}
loess.fit <- loess(df$lifeExp ~ df$year)
new.x <- seq(1953, 2006, length=100)
new.y <- predict(loess.fit, newdata=new.x)
lines(new.x, new.y, col="blue", size=5)
```

