---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## C++ with Rcpp

### Basics

For this section you will need the Rcpp and the microbenchmark packages.

```{r}
library(Rcpp)
library(microbenchmark)
```

You might also have to install Rtools from [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/) for Windows.

C++ is probably the most widely used general programming language today. Actually, about half the code of base R is written in C++! (the rest is about half and half R and Fortran)

Sometimes when you have some code that takes a while to run it is worthwhile to spend some time speeding it up. One way to do this is to rewrite part of the code in C++.

Say we have the following problem: we have a data set with points (x, y) and for each point we want to find the Euclidean distance to the origin $d=\sqrt{x^2+y^2}$. Here is a simple routine to do this:

```{r}
dist1 <- function(x, y) {
  n <- length(x)
  d <- rep(0, n)
  for(i in 1:n) d[i] <- sqrt(x[i]^2+y[i]^2)
  d
}
```

Let's see how long this takes:

```{r cache=TRUE}
x <- rnorm(1e6)
y <- rnorm(1e6)
microbenchmark(dist1(x, y))
```

Now of course we can immediately speed things up by vectorizing the routine:
 
```{r cache=TRUE}
dist2 <- function(x, y) {
  sqrt(x^2+y^2)
}
mb <- microbenchmark(dist1(x, y), dist2(x,y))
round(as.numeric(summary(mb)[, "median"]), 2)
```

Can we do even better?  

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericVector dist3(NumericVector x, NumericVector y) {
 int n=x.length();
 NumericVector dist(n);
 for (int i=0; i<n; ++i) dist[i]=sqrt(x[i]*x[i]+y[i]*y[i]);
 return dist;
}
```

```{r}
mb <- microbenchmark(dist1(x, y), dist2(x,y), dist3(x,y))
round(as.numeric(summary(mb)[, "median"]), 2)
```

Note: in the R markdown document the above chunk starts like this:

\```{r engine='Rcpp'}

this tells R markdown to treat this chunk as Rcpp code and compile it accordingly.

Actually, we can even do better than that:

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericVector dist4(NumericVector x, NumericVector y) {
  NumericVector dist;
  dist=sqrt(x*x+y*y);
  return dist;
}
```

```{r}
mb <- microbenchmark(dist3(x, y), dist4(x,y))
round(as.numeric(summary(mb)[, "median"]), 2)
```

Those of you who already know a bit of C++ are going to be quite amazed, because this is not even C++, it is sort of "vectorized" C++!

`r hl()$hr()`


To start let's discuss a few differences between R and C++ syntax:

-  in C++ every variable has to be explicitly defined.  
the most common data types are *int*, *double*, *char* and *bool*. 

-  in C++ (almost) every line ends with a ;  
-  vectors start with index 0, not 1  
-  the for loop is for(int i=0; i<n; ++i)  
-  the repeat loop is called a *do* loop and has the format

```{r eval=FALSE}
do {
# do stuff  
} while( condition );
```

- whereas R does a lot of type conversion, C++ does none. This can lead to occasional strange behavior:

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
double mydiv(int n, int m) {
  return n/m;
}
```

```{r}
mydiv(5, 2)
```

so the result is 2, not 2.5. The reason is that when we divide two integers, in C++ the result is always an integer.

To get the correct result we need to do the type conversion ourselves:


```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
double mydiv(int n, int m) {
  return n/double(m);
}
```

```{r}
mydiv(5, 2)
```


`r hl()$hr()`

To be linked to R the C++ routine has to start with with these lines:

```{r, eval=FALSE}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]

}
```


The easiest way to get started is to just use RStudio - File - New File - C++ File.


If the C++ routine takes as its argument a single value it can be defined in the usual C++ way, as in the *mydiv* routine. If we want to use a vector as an argument or the return object we need to use a special variable type called  NumericVector. It is just what it says it is.

### Debugging

Generally you should only turn fairly short code into C++, so debugging is not to big a problem. However, on occasion you might want to add a print statement to your code, so you can find out where it fails. here is how:

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericVector dist5(NumericVector x, NumericVector y) {
 int n=x.length();
 NumericVector dist(n);
 for(int i=0; i<n; ++i) {
   dist[i]=sqrt(x[i]*x[i]+y[i]*y[i]);
   Rcout<<i<<" "<<dist(i)<<"\n";
 }
 return dist;
}
```

```{r}
dist5(x[1:5], y[1:5])
```

### Sugar

Not only is Rcpp vectorized, many standard R functions have been ported to Rcpp. 

#### **Example**

We want to write a routine that simulates **Brownian motion** in $R^2$. That is, a stochastic process that moves as follows: if at time $t$ it is at $(x_0, y_0)$, then at time $t+\delta$ it is at $(x_0+\delta X, y_0+\delta Y)$ where $X,Y\sim N(0, 1)$.

Note: generating stochastic process can be quite slow in R because they are difficult to vectorize, with one step of a loop depending on the previous one.

The output of our function is going to be a nx2 matrix, so we will use the data type *NumericMatrix*.

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericMatrix bw(int n, double delta) {
 NumericMatrix xy(n, 2);
  xy(0, 0) = 0;
  xy(0, 1) = 0;
  for(int i=2; i<n; ++i) {
    xy(i, 0) = xy(i-1, 0) + delta*rnorm(1)[0];
    xy(i, 1) = xy(i-1, 1) + delta*rnorm(1)[0];
 }
 return xy;
}
```

```{r}
plot(bw(10000, 1), 
     type = "l", 
     xlab = "x", 
     ylab = "y", 
     col = "blue")
```

Notice the term *rnorm(1)[0]*, a little different from the standard R usage.

Anyone (like me!) who ever had to write a routine in C++ and needed a simple routine like rnorm which does not exist in C++ will find this very sweet! And that is why it is called sugar!


#### **Example (Fibonacci Numbers and the Golden Ratio)

let's write a routine that calculates the golden ratio via the Fibonacci numbers. First, these are defined by

$$
\begin{aligned}
&n_0    = 1\\
&n_1    = 1\\
&n_k    = n_{k-1}+n_{k-2}\\
\end{aligned}
$$
and the golden ratio is the limit

$$
\lim_{k \rightarrow \infty} \frac{n_k}{n_{k-1}}
$$
because of its definition the Fibonacci numbers are most easily calculated using recursion:

```{r}
fibR <- function(n) {
 if(n==0) return(0)
 if(n==1) return(1)
 return (fibR(n-1)+fibR(n-2))
}
fibR(10)
```

Let's write this with Rcpp:

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
int fib_cpp(const int n) {
 if(n==0) return(0);
 if(n==1) return(1);
 return fib_cpp(n-1)+fib_cpp(n-2);
}
```

```{r}
fib_cpp(10)
```

and now for the golden ratio:

```{r cache=TRUE}
golden_ratio <- function(n, fun) {
 fun(n)/fun(n-1)
}
mb <- microbenchmark(golden_ratio(10, fun=fibR), 
                     golden_ratio(10, fun=fib_cpp))
round(as.numeric(summary(mb)[, "median"]), 2)
```

not only is the cpp version much faster, in this example R is actually quite useless: while it does recursion, doing it to often quickly becomes a problem (because of memory issues). 

Now

```{r cache=TRUE}
golden_ratio(30, fun=fib_cpp)
```

The actual value of the golden ratio is of course $\frac{1+\sqrt{5}}2 =1.618..$.

### Using existing C++ routines. 

So far we used C++ to speed up calculations, and we could use Sugar to call standard R functions in the C++ routine. There is another use, though. C++ has been the main computing language in many fields for a long time, and so there exist a large number of excellent routines already written. Say you found one of these and want to use it in your R program. Here is how:

What we need to do is to write a Rcpp wrapper routine, that eventually calls the C++ subroutine. Here is an example:

```{r engine='Rcpp'}
# include <Rcpp.h>
using namespace Rcpp;

//function declaration
double sum_of_squares(double x[], int n);

// [[Rcpp::export]]
double sub_routine(NumericVector x) {
  int n=x.length();
  double y[n];
  double ssq;
  for(int i=0;i<n;++i) y[i]=x[i];
  ssq=sum_of_squares(y, n);
  return ssq;
}

double sum_of_squares (double x[], int n)
{
  double r=0.0;
  for(int i=0;i<n;++i) r+=x[i]*x[i];
  return r;
}
```

```{r}
sub_routine(1:10)
```

The *sum_of_squares* routine is pure C++, like any you might find on the web!

**Exercise**

Say we have a vector of numbers $(x_1,.,,x_n)$ and we want to find the k-means, that is $\frac1{k}\sum_{i=j}^{i=j+k-1} x_i$. Here is a routine in R:

```{r}
kmeans <- function(x, k){
  n <- length(x)
  y <- rep(0, n-k+1)
  for(i in 1:(n-k+1))
    y[i] <- mean(x[i:(i+k-1)])
  y
}
```

Rewrite the routine as C++ and compare the speeds on 

```{r}
set.seed(111)
x <- rnorm(1e5)
```

and k=5

Let's write this with Rcpp:

```{r engine='Rcpp', echo=FALSE}
# include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericVector kmeans_cpp(NumericVector x, int k) {
 int n=x.length();
 double tmp;  
 NumericVector y(n-k+1);
 for(int i=0; i<n-k+1; ++i) {
   tmp=0.0;
   for(int j=i;j<i+k;++j) tmp+=x[j];
   y[i]=tmp/k;
 } 
 return y;
}
```

Let's check first that the two routine do the same job:

```{r}
rbind(kmeans(x[1:10], 5),
kmeans_cpp(x[1:10], 5))
```

and now 

```{r}
mb <- microbenchmark(kmeans(x, 5), kmeans_cpp(x, 5))
round(as.numeric(summary(mb)[, "median"]), 2)
```

