---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## Interactive web applications with shiny

```{r}
library(shiny)
```

### Basic shiny apps

shiny is a package that has routines to create interactive web applications that run in a browser. 

As an example consider the app [Taylor Polynomials]( https://wolfgangrolke.shinyapps.io/taylor/).


Every app has two parts:

-  *ui* contains information about the layout of the page (**u**ser **i**nterface).   
-  *server* has the R routines that create the content of the page.

Let's begin by creating a page that does the following: it generates a data set with n observations from a standard normal distribution and then draws the histogram. The user can choose n:

#### **Example 1**

```{r example1, eval=FALSE}
ui <- fluidPage(
  numericInput(inputId = "n", # name of variable
               label = "Sample Size", # text on page
               value = 1000), # starting value
  plotOutput("plot1")
)
server <- function(input, output) {
  output$plot1 <- renderPlot({
    x <- rnorm(input$n)  
    bw <- diff(range(x))/50
    ggplot(data.frame(x=x), aes(x)) +
       geom_histogram(color = "black", 
                   fill = "white", 
                   binwidth = bw) + 
       labs(x = "x", 
            y = "Counts")
  })
}
shinyApp(ui, server)
```

If you are on a computer that has R running with shiny installed you can run this and all the apps here with

```{r eval=FALSE}
runGitHub(repo = "Computing-with-R", 
          username = "WolfgangRolke", 
          subdir = "shiny/example1")
```

you can also do this: File > New File > Shiny Web App, copy-paste commands into file, click Run App

Next we will improve the app in three ways:

-  nicer layout 
-  add the possibility to change the number of bins
-  add a title

#### **Example 2**

```{r example2, eval=FALSE}
ui <- fluidPage(
  titlePanel("Histogram App"),
  sidebarLayout(
    sidebarPanel( # Input widgets on left side of page
    numericInput(inputId = "n", 
               label = "Sample Size", 
               value = 1000, 
              width = "40%"), # size of box
    textInput(inputId = "nbins", 
              label = "Number of bins", 
              value = "50", 
              width = "20%") # size of box
    ),
    mainPanel(
      plotOutput("plot1")
    )  
))
server <- function(input, output) {
  output$plot1 <- renderPlot({
    x <- rnorm(input$n)  
    bw <- diff(range(x))/as.numeric(input$nbins)
    ggplot(data.frame(x=x), aes(x)) +
      geom_histogram(color = "black", 
                     fill = "white", 
                     binwidth = bw) + 
      labs(x = "x", 
           y = "Counts")
  })
}
shinyApp(ui, server)
```

notice also that instead of a numericInput I used a textInput, although nbins is meant to be a number. I often do this because I find the appearance of text inputs more pleasing. I can of course always turn it into a number with *as.numeric*.

Next we will use one of the most useful commands in shiny: *conditionalPanel*. With this we can control what happens in the app depending on the users choices.

Let's add the possibility to show a boxplot instead of the histogram. Notice that there are no bins in a boxplot, so we want to show the nbins input only when a histogram is done:

#### **Example 3**

```{r example3, eval=FALSE}
ui <- fluidPage(
  titlePanel("Histogram or Boxplot App"),
  sidebarLayout(
    sidebarPanel(
      numericInput("n", "Sample Size", 1000),
      selectInput("whichgraph", "What Graph?", 
          choices = c("Histogram", "Boxplot"),
          selected = "Boxplot"),
      conditionalPanel(
         condition = "input.whichgraph=='Histogram'",     
            textInput("nbins", 
                      "Number of bins", 
                      "50", 
                      width = "20%")
      )   
    ),
    mainPanel(
      conditionalPanel(
        condition="input.whichgraph=='Histogram'",
           plotOutput("plot1")
      ),
      conditionalPanel(
        condition="input.whichgraph=='Boxplot'",
           plotOutput("plot2")
      )   
    )  
))
server <- function(input, output) {
  output$plot1 <- renderPlot({
    x <- rnorm(input$n)  
    bw <- diff(range(x))/as.numeric(input$nbins)
    ggplot(data.frame(x=x), aes(x)) +
      geom_histogram(color = "black", 
                     fill = "white", 
                     binwidth = bw) + 
      labs(x = "x", 
           y = "Counts")
  })
  output$plot2 <- renderPlot({
    x <- rnorm(input$n)  
    df <- data.frame(x=x, z=rep(" ", length(x)))
    ggplot(df, aes(z, x)) + 
      geom_boxplot() + 
      xlab("")
  })
}
shinyApp(ui, server)
```

Another useful feature is the ability to create tabs. Let's say we want to show the histogram on one tab and the boxplot on another:

#### **Example 4**

```{r example4, eval=FALSE}
ui <- fluidPage(
  titlePanel("Histogram or Boxplot App"),
  sidebarLayout(
    sidebarPanel(
      numericInput("n", "Sample Size", 1000),
      conditionalPanel(condition="input.mytabs=='Histogram'",     
         textInput("nbins", "Number of bins", "50", width = "20%")
      )   
    ),
    mainPanel(
    tabsetPanel(
          tabPanel("Histogram", plotOutput("plot1")),
          tabPanel("Boxpot", plotOutput("plot2")),
          id="mytabs"
      )
    )  
))
```

Notice again the use of conditionalPanel: now the nbins box appears when the Histogram tab is selected.

Did you notice a slight flaw in the program? When we switch from one graph to the other the data is generated a new, so the graphs are not for the same data set. Let's fix this.

Also let's say we want on another tab to show some summary statistics:

#### **Example 5**

```{r example5, eval=FALSE}
ui <- fluidPage(
  titlePanel("Histogram or Boxplot App"),
  sidebarLayout(
    sidebarPanel(
      numericInput("n", "Sample Size", 1000),
      conditionalPanel(condition="input.mytabs=='Histogram'",     
         textInput("nbins", "Number of bins", "50", width = "20%")
      )   
    ),
    mainPanel(
    tabsetPanel(
          tabPanel("Histogram", plotOutput("plot1")),
          tabPanel("Boxpot", plotOutput("plot2")),
          tabPanel("Summary Statistics", uiOutput("text1")),
          id="mytabs"
      )
    )  
))
server <- function(input, output) {
  
  data <- reactive({
     rnorm(input$n)  
  })

  summaries <- reactive({
     x <- data()
     list(n = length(x), 
          xbar = round(mean(x), 3), 
          shat = round(sd(x), 3))
  })

  output$plot1 <- renderPlot({
    bw <- diff(range(data()))/as.numeric(input$nbins)
    ggplot(data.frame(x=data()), aes(x)) +
      geom_histogram(color = "black", 
                   fill = "white", 
                   binwidth = bw) + 
      labs(x="x", y="Counts")
  })
  
  output$plot2 <- renderPlot({
    x <- data()
    df <- data.frame(x=x, z=rep(" ", length(x)))
    ggplot(df, aes(z, x)) + 
      geom_boxplot() + 
      xlab("")
  })
  
  output$text1 <- renderText({
      lns <- "<table>"
      lns[2] <- paste("<tr><th>Sample Size&nbsp;</th><td>", summaries()[1], "</td></tr")
      lns[3] <- paste("<tr><th>Mean&nbsp;&nbsp;</th><td>", summaries()[2], "</td></tr")
      lns[4] <- paste("<tr><th>Standard Deviation&nbsp;&nbsp;&nbsp;</th><td>", summaries()[3], "</td></tr")
      lns[5] <- "</table>"
      lns
  })
}
shinyApp(ui, server)
```

This also shows that we can use html syntax in shiny. This is not a surprise because in the end it is a web page!

shiny apps are incredible versatile, you can write apps with a lot of stuff in it. You can even make little movies:

#### **Example 6**

```{r example6, eval=FALSE}
ui <- fluidPage(
  titlePanel("2-D Random Walk"),
  sliderInput("step","Step",
     min=1, max=100, value=1, step=1,
     animate=animationOptions(interval = 500, loop = FALSE)),
  plotOutput("plot", width="500", height="500")
)
server <- function(input, output) {
  data <- reactive({
    x <- cumsum(c(rep(0, 10), rnorm(1000)))  
    y <- cumsum(c(rep(0, 10), rnorm(1000)))
    data.frame(x=x, y=y)
  })

  make.plot <- reactive({
       xymax <- abs(max(data()))
       ggplot(data(), aes(x, y)) +
         lims(x=1.2*c(-xymax, xymax), y=1.2*c(-xymax, xymax)) +
         labs(x="x", y="y")  
  
  })

  output$plot <- renderPlot({
 
    for(i in 1:input$step) {
       plt <- make.plot() +
         geom_point(data=data()[10*i, ],
            aes(x,y), size=2, color="red") + 
         geom_line(data=data()[1:(10*i), ], aes(x,y), 
           size=0.25, color="blue", alpha=0.5)
    }     
    plt
  })
}
shinyApp(ui, server)
```

### Create/Run/Distribute shiny apps

There are a number of ways to create, run and distribute shiny apps:

1.  Create an app: 
   -  small apps are easiest done within RStudio, as in our example1  
   -  for larger apps you should have two separate files called ui.R and server.R is some folder, say myapp1. For example 1 they would look like this:
  
**ui.R**
```{r eval=FALSE}
shinyUI(fluidPage(
  numericInput("n", "Sample Size", 1000),
  plotOutput("plot1") 
))
```

**server.R**  
```{r eval=FALSE}
shinyServer(function(input, output) {
 output$plot1 <- renderPlot({
    x <- rnorm(input$n)  
    bw <- diff(range(x))/50
    ggplot(data.frame(x=x), aes(x)) +
    geom_histogram(color = "black", fill = "white", binwidth = bw) + 
    labs(x="x", y="Counts")
  })
})
```

then in the console run

```{r eval=FALSE}
runApp("PATH/myapp1")
```

where PATH is the folder path to myapp1.

2.  Distribute an app
   -  you can upload the folder myapp1 to github and then use
```{r eval=FALSE}
runGitHub("reponame", "username", 
          subdir = "shiny/myapp1")
```

here I  assume you have myapp1 in a folder called shiny. 

We will talk about Github a lot more soon.  

   -  you can make a zip file out of the folder myapp1, upload it to a web site and then run
   
```{r eval=FALSE}
runUrl("http://websiteurl/shiny/myapp1.zip")
```
   -  go to [https://www.shinyapps.io](https://www.shinyapps.io), set up a (free) account. Then you can upload a few apps. The big advantage of this is that even people who don't have R can now run your app. There are restrictions, though, on a free account, for example no more than 10 people can run an app at the same time.
   
Uploading an app to shinyapps should be very simple: run the app and click on the Publish button in the upper right corner. I have however had occasional problems with this method. If it does not work run the following command from the console:

```{r eval=FALSE}
shinyapps::deployApp("C:\\Users\\...\\name_of-app",
                     account="your_account_name")
```

For more information on how to get up and running with shinyapps.io read [https://shiny.rstudio.com/articles/shinyapps.html](https://shiny.rstudio.com/articles/shinyapps.html)
