---
title: Subsetting / Data Wrangling
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Vectors

Consider the following vector:

```{r echo=3}
x <- round(runif(10, 0, 10), 1)
names(x) <- LETTERS[1:10]
x
```

The elements of a vector are accessed with the bracket [ ] notation:

```{r}
x[3]
x[1:3]
x[c(1, 3, 8)]
x[-3]
x[-c(1, 2, 5)]
```

if a vector has names they can be used as well:

```{r}
x["C"]
x[c("A","D")]
```


There are also strange things one can do and sometimes get away with:

```{r error=TRUE}
x <- 1:10
names(x) <- letters[1:10]
x
x[0]
x[3.1]
x[3.6]
x[a]
```


Another way to subset a vector is with logical conditions:

```{r}
x[x > 4]
x[x>4 & x<7]
```

It is also possible to replace values in a vector this way:

```{r}
x[x<2] <- 0
x
```

This can be useful, for example to code a variable:

```{r}
Gender <- sample(c("Male", "Female"), 
                 size = 10, 
                 replace = TRUE)
Gender
GenderCode <- rep(0, length(Gender))
GenderCode[Gender=="Male"] <- 1
GenderCode
```

#Matrices and Data Frames

Consider the following data frame:

```{r, echo=6}
set.seed(111)
Age <- sample(18:25, size=10, replace=T)
GPA <- round(runif(10, 2, 4), 1)
Gender <- sample(c("Male", "Female"), size=10, replace=T)
students <- data.frame(Age, GPA, Gender)
students
```

Because a vector has rows and columns we now need to specify both:

```{r}
students[2, 3]
```

There are a variety of ways to do subsetting:

```{r}
students[, 1]
students[[1]]
students$Age
```

And yet another way to do this:

```{r eval=FALSE}
attach(students)
```
```{r}
Age
```

**Exercise**

What does this do?

```{r eval=FALSE}
x <- 1:10
x[]
```

`r hl()$hr()`

Although these seem to do the same there actually subtle differences. Consider this:

```{r}
students[, 1]
students[1]
students[[1]]
```

In the first and last case R returns a vector, in the second case a data frame with one column. 

It is possible to tell R not to do this type conversion in the first case

```{r}
students[, 1, drop=FALSE]
```

but this does not work for the [[1]] or \\$Age versions.

```{r}
students[1:3, 1]
students[-2, ]
students[1:4, -1]
students[Age>20, ]
```

You can have several conditions, put together with & (AND), | (OR) and ! (NOT), but some care is needed:

```{r}
students[Age>=20 & Age<=22, 1]
```

is fine but 

```{r eval=FALSE}
students[20 <= Age <= 22, 1]
```

does not work. 

#Lists

Subsetting of lists is very similar to data frames:

```{r}
mylist <- list(First=1:5, Second=LETTERS[1:8], Third=20:22)
mylist[1]
mylist[[1]]
mylist$Second
mylist[1:2]
mylist[[1:2]]
```

so [1] returns a list with just one element whereas [[1]] and $ do type conversion to a vector if possible. [1:2] yields the first two elements of the list.

The last one is strange, why is the result 2? Actually, it does this:

```{r}
mylist[[1]][2]
```

This can be quite confusing. Here is useful memory device 

*If list x is a train, x[[5]] is the content of car 5, whereas x[4:5] is the train consisting of cars 4 and 5*

#Partial Matching

There is one important difference between \$ and [[]], the first one does partial matching:

```{r error=TRUE}
x <- list(inside=1:3, outside=5:10)
x$o
x[[o]]
```

I don't however recommend to make use of this feature unless necessary. 

#Useful logic commands

```{r}
x<-1; y<-2; z<-3
c(x, y, z)>2.5
any(c(x, y, z)>2.5)
all(c(x, y, z)>2.5)
```

```{r}
x <- 1:3; y <- 1:3 
x == y
identical(x, y)
all.equal(x, y)
```

*identical* compares the internal representation of the data and returns TRUE if the objects are strictly identical, and FALSE otherwise.

*all.equal* compares the “near equality” of two objects, and returns TRUE or displays a summary of the differences. The latter function takes the approximation of the computing process into account when comparing numeric values. The comparison of numeric values on a computer is sometimes surprising!

```{r}
0.9 == (1 - 0.1)
identical(0.9, 1 - 0.1)
all.equal(0.9, 1 - 0.1)
```

but

```{r}
0.9 == (1.1 - 0.2)
identical(0.9, 1.1 - 0.2)
all.equal(0.9, 1.1 - 0.2)
```

How come $1.1-0.2 \ne 0.9$? This is because of machine precision issues:

```{r}
all.equal(0.9, 1.1 - 0.2, tolerance = 1e-16) 
```

#*subset*

Finally there is a command that was written for subsetting:

```{r}
subset(students, Age>20)
subset(students, Age>20 & Gender=="Male")
subset(students, Age>20, select = Gender)
subset(students, Age>20, select = Gender, drop=TRUE)
```

Notice that this last one results in a vector.

**Exercise**

The data set *upr* (part of *Resma3.RData*) has the application information provided to the University of all students that were eventually accepted between 2003 and 2013. Here are the first three students:

```{r}
head(upr, 3)
```

How many female students applied in either 2010 or 2011, had a highschool GPA of at least 3.0 and a freshman GPA between 3.0 and 3.5?

#*order*

if we need to sort a vector we have the sort command:

```{r}
sort(Age)
```

but sometime we want to sort one vector by the order of another:

```{r}
students[order(Age), ]
```

**Example** Lookup Tables

Say we want to write a function which many times need to calculate $\log(n!)$. We soon run into the following problem:

```{r error=TRUE}
log(factorial(175))
```

despite the fact that this is not a really big number (it is 732.3394). The problem is that internally R uses the gamma function to calculate factorials, and $\Gamma(175)$ is larger than what R can handle.

There is of course a simple solution:

$$
\log(n!) = \log(\prod_{i=1}^n i) = \sum_{i=1}^n \log i 
$$
but this turns out to be quite slow. Here is a better solution: we will make use of **Sterling's formal**:

$$
\begin{aligned}
&n! \sim n^ne^{-n}\sqrt{2\pi n} \\
&\log(n!) \sim n\log(n) - n+\frac12\log(2 \pi n) \\
\end{aligned}
$$

say we will need to find $\log(n!)$ for n ranging from 0 to 500, then we can create a lookup table. For small values of $n$ we use the exact formula, then the approximation:

```{r}
logfac <- 0:500
names(logfac) <- logfac
for(n in 0:50) 
  logfac[n+1] <- log(factorial(n))
for(n in 51:500) 
  logfac[n+1] <- n*log(n)-n+0.5*log(2*pi*n) 
```

and now we can find various values very easy:

```{r}
logfac[c("2", 5, 30, 301, 30)]
```

**Exercise**

why did I use 

logfac[c("2", 5, 30, 301, 30)]

and not

logfac[c(2, 5, 30, 301, 30)]
